COVID-19 Disparities in Quebec (Canada)

### 2023-05-02 Expérimentations (Théa)/2. Expérimentations (Bookdown)/Bookdown/Bookdown.Rmd
Multiple changes in Bookdown

---
title: "Réseaux sociaux et sentiment d’appartenance chez les immigrants établis au Québec"
author: "Vincent Bertrand-Huneaut, Jeanne Craig-Larouche, Théa Carlier, Noura Kone,  Antoine Martin"
date: "`r format(Sys.Date(), '%B %e, %Y')`"
present_to: "Vissého Adjiwanou"
type: "Rapport de recherche"
knit: bookdown::render_book
subtitle: Rapport de recherche
groupe_cours: "SOC3380, gr. 40"
site: bookdown::bookdown_site
bibliography: "library.biblatex"
csl: apalike-uqam.csl
editor_options:
  markdown:
    wrap: 80
    fontsize: 12pt
documentclass: article
link-citations: yes
institute: Université de Québec à Montréal
abstract: "Ce rapport de recherche présente les résultats d'une étude mixte visant à comprendre la relation entre l’usage des réseaux sociaux en ligne et l’intégration culturelle des personnes immigrantes de première génération établies au Québec depuis plus d'un an. Son objectif est de saisir comment les discours présents sur les réseaux sociaux en ligne, ainsi que l’entretien de réseaux diversifiés sur des sites comme Facebook, encouragent ou découragent les personnes immigrantes à s’intégrer dans la culture québécoise, et ainsi, à développer un sentiment d'appartenance pour le Québec. Pour répondre à ces questions, nous nous reposons sur la théorie de la fusion culturelle proposée par Stephen M. Croucher et Éric Kramer ainsi que sur les travaux de Friedrich Heckmann et de Wolfgang Bosswick sur l'intégration culturelle. Partant, nous avons conduit et analysé une série de 10 entretiens semi-directifs en parallèle d'une enquête par questionnaire (N = 65) auprès de personnes immigrantes de première génération établies à Montréal depuis plus d'un an. Mots clés : #Immigration #Intégration culturelle #Réseaux sociaux en ligne #Sentiment d'appartenance #Québec"
---

```{r setup, include=FALSE, warning=FALSE}
# Préférence knitr

knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo=FALSE)
```

@@

# Introduction

Au Québec, la pérennité de l'économie, de la démographie et de la langue
officielle repose en partie sur le renouvellement du flux migratoire
international. Dans ce contexte, l'intégration culturelle des personnes
immigrantes constitue un enjeu historique pour la province, dont le gouvernement
défend depuis les années 1960 la spécificité culturelle et politique au sein de
la fédération du Canada. Si l'intégration des personnes immigrantes est étudiée
depuis les années 1910 aux États-Unis, le récent développement des réseaux
sociaux en ligne offre depuis vingt ans une nouvelle manière d'appréhender ce
phénomène.

Dans l'état actuel de la littérature scientifique, nous savons que les réseaux
sociaux en ligne permettent aux personnes immigrantes de conserver un lien
affectif et culturel avec leur pays d'origine afin de réduire le stress
acculturatif généré par la précarisation liée au processus de migration. Dans le
même temps, les réseaux sociaux en ligne permettent aux personnes immigrantes
d'entretenir et d'accumuler du capital culturel et social mobilisable dans le
pays d'accueil.

Les études précédemment réalisées dans ce domaine reposent sur des séries
d'entretiens semi-directifs, ou sur une enquête par questionnaire. Dans le cas
des enquêtes qualitatives, la nature des échantillons tirés et le type
d'analyses réalisées ne permettent pas de quantifier la relation à l'étude ni de
généraliser les résultats à l'échelle de la population étudiée. Concernant les
enquêtes quantitatives, si l'enquête par questionnaire permet de mettre à
l'épreuve statistique la relation entre l'usage des réseaux sociaux en ligne et
l'intégration culturelle des personnes immigrantes, celle-ci peine à fournir des
explications précises quant aux modalités du phénomène étudié.

Dans ce contexte, nous avons réalisé une étude mixte regroupant 10 entretiens
semi-directifs ainsi qu'une enquête par questionnaire (N = 65) visant à
reproduire en contexte québécois les études proposées par la littérature
scientifique. Notre objectif est de comprendre la relation entre l'usage des
réseaux sociaux en ligne et l'intégration culturelle des personnes immigrantes
établies au Québec. Plus précisément, nous nous intéressons au sentiment
d'appartenance que peuvent développer les personnes immigrantes à travers leur
utilisation des réseaux sociaux en ligne.

@@

<!--chapter:end:01-intro.Rmd-->

@@

# Revue de littérature

## Intégration culturelle

Pour comprendre ce que signifie le sentiment d'appartenance des personnes
immigrées établies au Québec, il est nécessaire de se pencher sur le lien qui
unit ces dernières à leur pays d'accueil.

Cette relation intéresse les sociologues depuis le début du XIX^e^ siècle. On en
trouve d'abord la trace aux États-Unis, au sein de la première école de Chicago.
Les travaux de William I. Thomas [-@thomasPolishPeasantEurope1918] et Robert E.
Park [-@parkHumanMigrationMarginal1928] contribuent à définir un premier concept
servant à appréhender ce rapport : l'assimilation. Celle-ci se définie comme : «
un processus individuel de convergence des caractéristiques des [personnes
immigrées] vers les caractéristiques moyennes de la société d'accueil »
[@safiPenserIntegrationImmigres2011, p. 150].

Après avoir dominé durant le XXe siècle, le concept d'assimilation est remis en
question. Friedrich Heckmann remarque que celui-ci est normatif, en ce qu'il
prend pour acquis que la culture du pays d'accueil forme un ensemble homogène
[-@heckmannIntegrationIntegrationPolicies2005, p. 12].

S'inspirant de la théorie de l'intégration des systèmes publiée par David
Lockwood en 1964, Hartmut Esser propose d'appréhender ce lien comme une seconde
socialisation possédant quatre dimensions : 1) la capacité des personnes
immigrantes à mobiliser les institutions de leur pays d'accueil ; 2) leur
placement dans le marché du travail, de la vie conjugale, etc. ; 3) leur
capacité à interagir avec d'autres personnes dans la société d'accueil ; et 4)
le sentiment d'appartenance qu'elles développent envers leur pays d'accueil ou
d'autres sous-ensembles lui appartenant [-@esserDoesNewImmigration2004, p.
1131].

Or, cette nouvelle conceptualisation du lien entre les personnes immigrées et
leur pays d'accueil se heurte à la critique de Young Kim, qui avance que
celui-ci ne constitue pas simplement une deuxième socialisation, mais implique
également une désocialisation plus ou moins prononcée vis-à-vis de la diversité
culturelle présente dans le pays d'origine
[-@kimBecomingInterculturalIntegrative2001, p. 53]. Quelques années plus tard,
Stephen M. Croucher et Éric Kramer développent la théorie de la fusion
culturelle. A l'instar des travaux de Kim, la fusion culturelle se comprend
comme « [le] mélange entre l'apprentissage de nouveaux comportements et le
maintien [relatif] d'anciens comportements »
[-@croucherCulturalFusionTheory2017, p. 97]. Néanmoins, Croucher et Kramer
introduisent une nouvelle dimension dans leur concept, en avançant que la fusion
culturelle est simultanée. En d'autres termes, cela signifie que la culture des
personnes immigrantes fusionne avec la diversité culturelle présente dans le
pays d'accueil pour produire une forme altérée de la diversité culturelle
initiale [-@croucherCulturalFusionTheory2017, p. 103].

## Réseaux sociaux en ligne et intégration culturelle

La migration constitue une rupture dans le processus de socialisation car elle
déracine les personnes migrantes de leur environnement. Dans ce contexte, elles
expérimentent ce que des chercheurs en psychologie comme John W. Berry et al.
[-@berryImmigrationAcculturationAdaptation1997] appellent du « stress
acculturatif ». Celui-ci est caractérisé par de possibles dégradations de la
santé comme l'anxiété, la dépression, la confusion identitaire ou encore un
sentiment de marginalisation de la part des personnes établies dans le pays
d'accueil. Selon Berry, ce stress et cette marginalisation (effectifs et/ou
perçus) sont causés par les différences culturelles qui oppose le pays d'origine
au pays d'accueil.

L'étude d'Eric Forbush et de Brooke Foucault-Wells [-@forbushSocialMediaUse2016]
sur l'usage des médias sociaux par des étudiants chinois aux États-Unis, ainsi
que l'étude d'Amanda Alencar [-@alencarRefugeeIntegrationSocial2017] parmi les
réfugiés syriens établis aux Pays-Bas nous apprennent que les personnes
immigrantes utilisent régulièrement des réseaux sociaux en ligne comme Facebook,
mais aussi des applications de messagerie instantanée comme WhatsApp pour
communiquer avec leurs proches restés au pays. Dans une étude sur l'intégration
culturelle des étudiants chinois établis aux États-Unis, Shaoke Zhang
[-@zhangBabelGreatWall2012] constate que les réseaux sociaux en ligne permettent
de réduire le stress acculturatif généré par la migration, de diminuer
l'incertitude quant à la situation des personnes immigrantes et de réduire le
fossé les séparant de la culture du pays d'accueil. Dans une étude menée en
Nouvelle-Zélande par Kingsley T. Ihejirika et Maja Krtalic
[-@ihejirikaMovingMediaExploration2021], on peut lire que les personnes
immigrantes utilisent les réseaux sociaux en ligne pour s'intégrer en
s'imprégnant des normes, de la langue et du mode de vie néo-zélandais afin de
pouvoir interagir avec la population et les institutions locales. Leurs travaux
révèlent que les personnes immigrantes utilisent les réseaux sociaux en ligne
pour collecter des informations, communiquer avec leur famille restée au pays,
construire de nouvelles relations, réseauter et apprendre la langue du pays
d'accueil. La proportion de ces usages varie en fonction des caractéristiques
sociodémographiques des migrants, mais aussi du contexte socioéconomique. En
reprenant des matériaux issus de l'enquête ALTERECOLE, Évelyne Barthou
[-@barthouMigrationSansRupture2020] montre que les réseaux sociaux en ligne
facilitent également l'émergence d'identités plurielles et d'une culture
composite chez les immigrants marocains établis en France.

Or, ces résultats sont pondérés par le travail de Lee Komito
[-@komitoSocialMediaMigration2011] dans son étude sur les polonais.es et les
philippin.es établies en Irlande. Le chercheur remarque que l'usage des réseaux
sociaux en ligne peut contribuer à isoler les personnes immigrantes en leur
permettant de rester connecté à la culture de leur pays d'origine. En outre,
dans leur travaux sur les jeunes immigrant.es établis au Québec, Nicole Gallant
et Céline Friche relèvent que si les médias sociaux en ligne facilitent
l'entretient de réseaux sociaux et d'appartenances préexistantes, ceux-ci : «
semblent avoir peu d'incidence sur le développement d'un réseau social dans le
milieu d'accueil, si ce n'est pour colorer les rapports au quotidien »
[-@gallantEtreIciLabas2010, pp.64].

## Le place du discours sur les réseaux sociaux en ligne

La production et la diffusion d'informations en ligne diffèrent de celle des
médias traditionnels, car internet permet une interaction instantanée entre le
producteur et le consommateur de contenu. Comme le démontrent André Mondoux et
Marc Ménard dans leur livre *Big data & Société* [-@mondouxBigDataSociete2018],
la logique algorithmique place les utilisatrices et les utilisateurs des réseaux
sociaux en ligne dans une relation circulaire entre leur consommation et des
propositions ciblées.

Dans un chapitre d'ouvrage collectif, Ehsan Dehghan et Axel Bruns révèlent que
la racialisation discursive observée sur les réseaux sociaux en ligne à
l'endroit des personnes immigrantes n'est pas tant dû à la structure des
plateformes qu'au comportement stratégique d'utilisateurs souhaitant promouvoir
leurs idées [-@dehghanDynamicsPolarisationAustralian2022, p.64‑65]. Dans le même
temps, des études comme celles de Jeroen Hopster ou encore Manoel Horta Ribeiro
et al. démontrent comment la logique algorithmique des réseaux sociaux en ligne
facilite le développement de communautés véhiculant des discours populistes -
souvent xénophobes et racistes - par l'entremise de la suggestion automatique de
contenu et le système de commentaires sur YouTube
[-@hopsterMutualAffordancesDynamics2021, p. 559;
-@ribeiroAuditingRadicalizationPathways2020, p. 140]. James P. Walsh parle même
de « nativisme digital » pour désigner l'usage stratégique de Twitter dont font
preuve certains groupes d'extrême droite [-@walshDigitalNativismTwitter2021, p.
5]. Selon lui, durant la campagne électorale canadienne de 2019, les messages
racistes à l'attention des personnes immigrantes sur Twitter ont généré un
engouement inédit, exacerbant les discours anti-immigration. Son travail
soutient que le nativisme - une idéologie xénophobe s'opposant à la migration et
reposant sur le principe de racialisation - est en train d'être restructuré par
les réseaux sociaux en ligne. Selon Walsh, les groupes nativistes ont
instrumentalisé les possibilités offertes par Twitter afin de véhiculer un
discours raciste et anti-immigration parmi la population et les institutions
canadiennes [@walshDigitalNativismTwitter2021, p. 19].

@@

<!--chapter:end:02-revue-litterature.Rmd-->

@@

# Cadre théorique

## Intégration culturelle

Dans le cadre de notre étude, nous avons produit une définition de l'intégration
culturelle à partir des travaux de Heckman, Bosswick
[-@heckmannIntegrationMigrantsContribution2006, p.10], Kim
[-@kimBecomingInterculturalIntegrative2001, p.53], Cramer et Kroucher
[-@croucherCulturalFusionTheory2017, p. 97]. Selon ces auteurs, le processus
d'intégration culturelle implique, à l'échelle microsociologique (l'individu) et
mésosociologique (la population des personnes immigrantes de première
génération), une transformation de la diversité culturelle véhiculée par les
personnes immigrantes lorsqu'elles s'établissent dans un nouveau pays. Dans le
même temps, à un niveau macrosociologique (les populations immigrantes établies
depuis plusieurs générations), la diversité culturelle du pays d'accueil est
transformée au contact des populations immigrantes nouvellement arrivées.

Dans ce contexte, l'identité nationale se fonde sur une construction sociale
historiquement située, dont les modalités se transforment continuellement au fil
du flux migratoire (y compris lorsque celui-ci prend la forme d'une invasion ou
d'une colonisation). Partant, à l'échelle micro et méso, le développement d'un
sentiment d'appartenance national repose sur la proximité entre la diversité
culturelle véhiculée par les individus et celle du pays dans lequel ils se
trouvent.

Pour nous aider dans l'élaboration d'un modèle d'analyse, les travaux de Marcin
Gońda et al. [@gondaMeasuringCulturalDimension2021, p.245-246] proposent de
diviser l'intégration culturelle en trois dimensions : 1) la langue ; 2) les
valeurs, normes et pratiques ; et 3) l'identification culturelle. Partant Gońda
et al. proposent de mesurer ses trois dimensions à l'aide d'indicateurs comme :
1) la connaissance et la compréhension des expressions et/ou de l'accent du pays
d'accueil ; 2) le type et la fréquence de consommation de contenu culturel du
pays d'accueil ; et 3) le sentiment d'appartenance au pays d'accueil.

Enfin, les travaux de Loïc Wacquant sur le processus de racialisation
[-@wacquantResolvingTroubleRace2022] révèlent que certaines populations comme
les Afro-Américains peuvent être graduellement catégorisées et stigmatisées sur
la base de rapports de pouvoir dérivés d'objectifs économiques, comme
l'esclavage aux Etats-Unis.

## Réseaux sociaux en ligne

Pour mieux cerner la notion de réseaux sociaux en ligne, nous avons repris les
travaux de Andreas M. Kaplan et Michael Haenlein [-@kaplanUsersWorldUnite2010,
p60-62]. Selon ces auteurs, les réseaux sociaux en ligne peuvent être divisés en
quatre types :

```{r}
local({
  .Table <- data.frame (
    reseaux  = c(
      "**Partage de médias**",
      "**Blogs**",
      "**Communautés en ligne**",
      "**Réseaux sociaux**"
    ),
    description = c(
      "Permet à des utilisateurs de produire, trier et partager des médias avec d'autres personnes ou groupes",
      "Permet aux utilisateurs de publier en ligne", 
      "Permet aux utilisateurs de partager des informations spécifiques à travers des forums",
      "Permet aux utilisateurs de construire des réseaux sociaux amicaux et professionnels"
    ),
    exemple = c("YouTube", "Twitter", "Reddit", "Facebook")
  )
  pander(.Table, split.cells = c(20, 40, 10), split.table = Inf,
         caption = "Typologie des réseaux sociaux",
         col.names = c("", "Descriptions", "Exemple")
  )
})
```

## Cadre conceptuel

Le schéma suivant cette page illustre, au regard de nos choix théoriques, notre
approche concernant l'étude de la relation entre l'utilisation des réseaux
sociaux en ligne et l'intégration culturelle des personnes immigrantes établies
au Québec.

\setcounter{figure}{`r paste(no_graphic <- no_graphic+1)`}
![Cadre conceptuel schématisé](images/figure_6.png){width=80%}

## Hypothèses de recherche

### H1 - Contact interculturel et sentiment d'appartenance

>>>>

---
title: "COVID-19 Disparities in Quebec (Canada)"
author: "V. Adjiwanou, G. NGalé, S. Lardoux, N. Zahedinameghi"
date: "`r format(Sys.Date(), '%B %e, %Y')`"
present_to: "Vissého Adjiwanou"
type: "Article"
knit: bookdown::render_book
subtitle: A Comparative Analysis of Exposure Risks, Consequences, and Access to Resources among Immigrants and Natives
site: bookdown::bookdown_site
bibliography: "library.biblatex"
csl: apalike-uqam.csl
editor_options:
  markdown:
    wrap: 80
    fontsize: 12pt
documentclass: article
link-citations: yes
institute: Université du Québec À Montréal and Université De Montréal
abstract: "The immigrant population is one of the most vulnerable in Canada, with a relatively high risk of unemployment, a low social support system, and linguistic barriers. However, their experiences with public health interventions during the COVID-19 pandemic have not been adequately explored. The purpose of this article is to analyze Covid-19 exposure risk and its family implications for both natives and immigrants in the province of Quebec. Additionally, we analyze immigrants' access to government aid and compare it to their native-born counterparts. The study consists of survey data collected from 1,000 immigrants and 500 non-immigrants Quebecers between August and November 2020. Our descriptive and exploratory analysis reveals significant differences between the two groups regarding exposure risk, family consequences, and access to resources. We conclude that the current pandemic exacerbates health disparities in the province, with particularly dire repercussions for immigrants' families. We also discuss several socio-cultural strategies to mitigate inequality and recommend policies to address the needs of vulnerable populations."
---

@@

# Introduction / background

The COVID-19 pandemic has highlighted the vast ethnic, social, economic, and
health disparities in high-income countries, including Canada. One of the most
vulnerable segments of these affluent societies is the immigrant population,
that has become widely diversified with more and more people coming from
developing countries, and for whom integration into their new society may be
more difficult. The coronavirus pandemic brings a new challenge for these
immigrant populations, directly or indirectly through the tensions between the
ethnic groups that it helps to generate [@coates2020Covid19]. These immigrants
face language barriers, low social capital, housing insecurity, and severe
poverty, rendering them particularly susceptible to the effects of the current
pandemic. Immigrants are also more likely to be employed as essential workers
where social distancing, sanitary precautions, and self-isolation, measures put
in place to mitigate the pandemy, are challenging [@reid2021Migrant;
@yi2021Health; @zhang2021COVID19]. More importantly, migrants may encounter
various forms of discrimination, leading to social exclusion and restricted
access to healthcare and social services [@spiritus-beerden2021Mental]. It is
therefore necessary to understand how this pandemic affects these immigrant
populations differently and the native populations in Quebec. In doing so, this
article will reveal a better understanding of how the current pandemic
aggravates existing inequalities and challenges in the province for immigrants
and their descendants.

<!-- The province of Quebec has one of the highest concentrations of immigrants in Canada, including permanent residents, temporary residents, and refugees [@governmentofcanada2017Daily; @immigration2005Recent]. In 2016, 1.2 million immigrants were living in Quebec, an increase from 8% in 1990 to 13.7% of the population in 2016 (Institut de la statistique du Québec, 2020). <!--As of 2019, the increase in net new residents (60,700 versus 42,200 in 2018) and lower interprovincial migration losses (-1,500 versus -6,100 in 2018) contributed to Quebec's population growth. -->

Since the start of the pandemic, most of the confirmed cases were identified in
Montreal and areas with high immigrant populations [@miconi2021Ethnocultural;
@shields2020COVID19]. The coronavirus pandemic has infected nearly 450,000
people in Quebec and has resulted in more than 11,000 deaths
(<https://www.quebec.ca/sante/problemes-de-sante/a-z/coronavirus-2019/situation-coronavirus-quebec>,
consulted on octobre 29, 2021). This makes it the worst affected province in
Canada and one of the worst affected regions in the world during the first wave
of the pandemic. Although the majority of these deaths took place in elderly's
homes, and therefore affected more this segment of the population, its overall
influence on individuals is not to be overlooked. It has been shown that, on the
one hand, immigrants and their families have a higher risk of being exposed to
the pathogen, and, at the same time, they are more vulnerable to racism and
discrimination and face barriers to accessing social support [@2021CRARR] and
information, especially for those who didn't speak or understand the two
official langages.
<!--In addition to the direct consequences of the pandemic, the measures taken by various governments around the world are also not without consequences on families. -->Several
studies have thus shown excess mortality in ethnocultural minority populations
[@bertocchi2020Covid19], higher unemployment, greater loss of income and
negative consequences for children and families in several countries. This
differentiated effect can be explained in different ways. On the one hand, in
terms of the direct consequences of the pandemic, the situation of
precariousness and greater poverty among immigrants exposes them more to the
more harmful consequences of the disease on their state of health. Likewise,
immigrants often work in jobs where social distancing is difficult to enforce.
On the other hand, indirectly, the solutions to the crisis affect immigrants and
natives differently. The complete shutdown of tye economy with its set of
adjustments has greatly impeded the employability of immigrants, especially
newcomers who have not yet grasped the workings of society and who cannot
benefit from state subsidies. But, it is also shown that these effects are not
uniform on the whole of the immigrant population, itself, very diverse.

Despite the growing focus on the differential impacts of the COVID-19 on
immigrants and non-immigrants segment of the population, there is still very
little knowledge about how the pandemic and its related public policy impact the
lives of immigrants and their families. The existing health datasets (largely
used to assess this effect) are not necessarily representative of the immigrant
population and do not offer an accurate picture of the complex social status and
health outcomes among this group [@choi2021Studying; @gagnon2021Immigration;
@khan2015Visible]. In most cases, these datasets lack data on the residence
length of immigrants and combine newcomers with long-term settlers
[@edge2013Discrimination; @hayward2021Clinical]. As a result, they may not allow
a detailed assessment of the nature and extent of risk exposure in sub-segments
of populations, as well as whether interventions are reaching those most at
risk. In addition, the family-level impacts of the pandemic are still poorly
understood. While a body of research has explored the association between the
pandemic and its potential effects on intimate partner violence and marital
dissolution [@bhatnagar2021Navigating; @koshan2020COVID19; @morrison2021Chapter;
@prime2020Risk], few studies have examined negative familial repercussions in
the context of immigration status. There is ample evidence that sustained family
relationships greatly enhance immigrants' lives on several fronts, including
fostering economic prosperity [@marger2001Sociala], psychological well-being
[@beiser2002Povertya], social network development [@evansluong2017Familya], and
educational success for children [@munroe-blum1989Immigranta]. The family is,
therefore, a key component of long-term social mobility and economic success of
immigrants.

<!--Faced with the pandemic, the governments of Canada and Quebec have implemented a number of measures to reduce its impact. These policies can be broken down at two levels, direct support to families, including the Canadian Emergency Benefit, which can provide employees and self-employed workers affected by Covid-19 with a financial benefit of up to $ 2,000. On the other hand, the government supports many companies to keep their employees or supports them directly when the revenues from their activities have fallen considerably. While these measures allow families to cope better with the crisis, the fact remains that the changes brought about by this crisis can affect other dimensions of life. How these measures make a difference in the face of the pandemic remains to be seen. -->

Based on an original biographic survey collected around the first wave of the
pandemic in august-november 2020 on immigrants and the native population of
Québec, this study mainly aims to understand the family exposure, consequences,
and responses of the pandemic on immigrant and native Quebec populations. We
assess a whole range of different measures of each outcome. We divided the
consequences in three dimensions, 1) access to ressources, 2) ability to perform
duties and 3) employment and income. We seek also to analyse in what ways these
outcomes differ according to the immigrants length of residence. By doing so,
this study provide a clear and global picture of the situation of different
population groups in Quebec in the face of the COVID-19.
<!-- To respond to these questions, we will use data collected around the first wave of the pandemic in august-november 2020. Our study aims to fill the gap by analyzing group-specific differences in exposure risk, family consequences, and access to resources among immigrants and Quebecers. We also seek to shed light on several socio-cultural strategies to mitigate inequality and recommend relevant policy solutions. The detailed findings will help policymakers address health inequalities and improve equity in Quebec while developing adequate measures and programs. -->

<!-- Introduction  -->

<!-- Like the rest of Canada, Quebec has experienced a substantial increase in its immigration in recent years. The immigrant population in the province thus increased from 8% in the 1990s to 13.7% of the population, or 1.2 million individuals in 2016 (Institut de la statistique du Québec, 2020). This immigration is also unique in that it has become widely diversified with more and more people coming from developing countries, and for whom integration into their new society may be more difficult. The coronavirus pandemic brings a new challenge for these immigrant populations, directly or indirectly through the tensions between the ethnic groups that it helps to generate [@coates2020Covid19]. It is therefore necessary to understand how this pandemic affects these immigrant populations differently and the native populations in Quebec. -->

<!-- The coronavirus pandemic has affected nearly 100,000 people in Quebec and has resulted in more than 6,000 deaths. This makes it the worst affected province in Canada and one of the worst affected regions in the world. Although the majority of these deaths took place in homes for the elderly, and therefore affected this segment of the population more, its overall influence on individuals is not to be overlooked. In fact, in addition to the direct consequences of the pandemic, the measures taken by various governments around the world are also not without consequences on families. Several studies have thus shown excess mortality in ethnocultural minority populations [@bertocchi2020Covid19], higher unemployment, greater loss of income and negative consequences for children and families in several countries. -->

<!-- This differentiated effect can be explained in different ways. On the one hand, in terms of the direct consequences of the pandemic, the situation of precariousness and greater poverty among immigrants exposes them more to the more harmful consequences of the disease on their state of health. Likewise, immigrants often work in jobs where social distancing is difficult to enforce. On the other hand, indirectly, the solutions to the crisis affect immigrants and natives differently. The complete shutdown of the company with its set of adjustments has greatly impeded the employability of immigrants, especially newcomers who have not yet grasped the workings of society and who cannot benefit from state subsidies. But, it is also shown that these effects are not uniform on the whole of the immigrant population, itself, very diverse. -->

<!-- Faced with the pandemic, the governments of Canada and Quebec have implemented a number of measures to reduce the impact of the pandemic. These policies can be broken down at two levels, direct support to families, including the Canadian Emergency Benefit, which can provide employees and self-employed workers affected by Covid-19 with a financial benefit of up to $ 2,000. On the other hand, the government supports many companies to keep their employees or supports them directly when the revenues from their activities have fallen considerably. While these measures allow families to cope better with the crisis, the fact remains that the changes brought about by this crisis can affect other dimensions of life. How these measures make a difference in the face of the pandemic remains to be seen. -->

<!-- This study mainly aims to understand the family consequences of the pandemic on immigrant and native Quebec populations. Therefore, we will seek to analyse in what ways its consequences differ according to the region of origin of immigrants, length of residence, age and sex. -->

@@

<!--chapter:end:01-introduction.Rmd-->

@@

# Current body of literature and knowledge gaps

The literature on the racial and ethnic health disparities during the COVID-19
crisis can be divided into two main categories: studies addressing the
socio-economic determinants of health and those examining the delivery of health
interventions and policies. The first category is primarily comprised of
research that focuses on the link between socio-economic status (SES) of
high-risk minority groups and the differential impact of the pandemic
[@drefahl2020Populationbased; @greenaway2020COVID19; @lopez2021Racial;
@pacheco2020Job; @sundaram2020Individual]. For instance, an inter-country study
has found that Middle Eastern, African, and Nordic immigrants in Sweden have a
higher risk of being hospitalized and even dying from COVID-19 when compared to
the Swedish born population. In the same study, the SES, proportion of
working-age households, and neighborhood density were significantly associated
with the mortality rate [@rostila2021Disparities]. This trend of research
demonstrates that racial and ethnic minorities are more likely to experience
stress factors, not only due to the hazards of infection but also because of
economic and social marginalization [@alahmad2020COVID19; @choi2020Buffering;
@endale2020COVID19; @fitzpatrick2020Fear; @serafini2021Psychological]. The
intersectionality of race, sex, and health conditions has also been explored in
the Canadian context of the pandemic. @ingen2021Neighbourhoodlevel demonstrated
that Ontarian neighborhoods with immigrant and racialized populations have a
disproportionate share of COVID-19 infections. In addition, areas with high SES
report a lesser number of COVID cases than low SES ones
[@ingen2021Neighbourhoodlevel]. Another study conducted in Quebec confirms the
strong association between immigrants' backgrounds, SES, and COVID-19 related
psychological stress [@miconi2021Ethnocultural]. The most recent reviews on the
topic suggest that immigrants disproportionately live in crowded and inadequate
housing and are employed in essential industries, leaving them with severe
income insecurities and poor physiological/psychological well-being
[@chew2020Clinical; @davis2020Impact]. Studies on SES and health disparities
have emphasized that this situation is further complicated for immigrant women,
constrained by gender and afflicted by a significant power imbalance in the
household and workplace [@alarcao2021Intersections;
@esegbona-adeigbe2020COVID19; @jehn2021COVID19; @mo2020Differential;
@nardon2021Skilled; @sabri2020Effect; @st-denis2020Sociodemographic]. According
to Lightman, immigrants and women of color are overrepresented in long-term care
facilities in Canada, where COVID infection rates have been exceptionally high.
They are also subject to exclusionary social and economic practices
[@lightman2021Caring].

The second category of the literature is primarily comprised of policy analyses,
research on public health communications, examining the efficacy of COVID
initiatives, the optimal allocation of government aids, and ethical issues
associated with applying new health interventions to vulnerable populations and
their communities [@aragona2020Negative; @cross2021Coronavirus;
@desai2020COVID19; @doyle2020Migrant; @gilman2020Modelling;
@hargreaves2020Targeting; @langellier2020Policy; @nezafatmaldonado2020Engaging;
@silverman2020Ethics; @wilson2020Ethics]. Some of the questions that this body
of research addresses include: how best to respond to COVID-19 in immigrant
communities? How do public health policies impact the socio-economic status and
psychological well-being of different sub-populations? What is the structure of
the planning and implementation of government aid programs during the current
crisis? And, how can we design public health communication strategies in a way
that is culturally sensitive and inclusive?

There are several knowledge gaps in the existing literature. Most of the cited
studies do not include household data and family structure of immigrants as
variables that affect exposure risks and health outcomes of COVID-19.
Consequently, the relationship between immigrant status and COVID-19 is not
accounted for by household size, and family characteristics (i.e., income,
education, religion). A key reason is that most of the health data used in such
research projects are acquired from national registers or administrative
databases, which provide little information on immigration status and kinship.
However, as the study of Thomeer and her colleagues theorizes, the extent of
adverse social and economic impacts of COVID-19 is significantly conditional on
the composition of an immigrant's family [@thomeer2020How]. Households that
include a spouse/partner, children, or multigenerational family members most
likely experience a different level of social, psychological, and medical
vulnerability than single or nuclear ones. Therefore, studies of immigrant
populations should account for the breakdown of family configuration to better
understand the mechanisms linking social determinants and health outcomes.
Another knowledge gap concerns the current state of Canadian studies of COVID-19
and its public health crisis. There is a dearth of literature on the challenges
faced by immigrants and their families in the country. Although the burden of
COVID-19 infection is disproportionately higher among these marginalized groups
(St-Denis, 2020; Sundaram et al., 2020), research on the issues they are
experiencing is sparse and often limited to one aspect.
<!--Likewise, there is limited focus on the effectiveness of COVID-19 crisis management for immigrants at the policy level. In response to the mounting cases of COVID, the federal government and provinces and territories of Canada have launched multi-pronged public health initiatives to protect society from the epidemic (Migone, 2020). However, many open questions remain unanswered, including, do interventions efficiently reach the migrant groups? What are the likely effects of each program on different immigrant communities? How can an intervention be tailored to the needs of a particular migrant group and their families? -->

@@

<!--chapter:end:02-body-of-litterature.Rmd-->

@@

# Theoretical framework

Family as a unit of analysis can be particularly fruitful when studying health
inequalities in the context of a societal-wide crisis. Through the lens of the
COVID-19 pandemic, the following points illustrate why exploring family dynamics
provides additional insight into the underlying mechanisms of health
disparities. First, the family is the most fundamental unit of social
organization. It facilitates access to resources and the transmission of social
and psychological capital from one generation to the next. Therefore, a better
analysis of how crises and adverse life events affect people requires a better
understanding of how families relate to each other and interact with external
forces. Second, during public health emergencies, families may be thrown into
different structures due to unprecedented financial, mental, and physical
challenges. The risk of family dissolution, for example, can be heightened when
members must spend unusually long periods together at home and have limited
opportunities to engage with friends and the wider community. Conversely,
families can develop closer bonds through self-sacrifice, shared
responsibilities, and increased care and support. As the dependency status of
seniors increases, the elderlies are also more likely to be integrated into
formerly nucleus families. We posit that the extent to which a family can adapt
its functions to meet new challenges is a predictor of the well-being of its
members and the resilience of its broader population. Third, the family is a
crucial link to social institutions, such as the health care system, schools,
and government agencies whose decisions have significant downstream impacts.
Families are viewed by many public-level entities as a reference when allocating
resources, enforcing new standards, and responding to emerging needs and
demands. From this perspective, the family operates at the nexus of individual
and societal structures, embedding the tripple effects of public policy in its
most consequential ways. Finally, the family has a crucial role to play in
determining the life course of its members. As COVID-19 elevates the importance
of familial bonds, the post-pandemic period might be marked by an increased
influence of families on the lives of their members and the social structures
they inhabit.

The above conceptualization echoes several other approaches to studying
long-term health inequalities. For instance, life course theory emphasizes the
significance of family ties and social environment as they structure prospects
for individuals across their life span (Elder, 1977; Elder Jr., 1998; Macmillan,
2005; Settersten et al., 2020). Likewise, social capital theory and the
stress-buffering hypothesis consider family systems a crucial factor in health
and well-being (Choi and Jun, 2020; Cohen and McKay, 1984; Cohen and Wills,
1985; Harpham et al., 2004; Kawachi and Berkman, 2001; Lavee et al., 1987;
Mandelbaum et al., 2020). They show that the family can offer tangible (e.g.,
food, shelter, and money) and intangible resources (e.g., support, social
status, psychological comfort, and institutional advantages), protecting its
members from adverse life events. Similarly, the social-ecological model and the
theory of fundamental causes highlight how the SES, familial and extrafamilial
factors pattern people's life trajectories and health outcomes (Barkan and
Rocque, 2018; Clouston and Link, 2021; Figge et al., 2018; Kim, 2020; Libman et
al., 2012; Ungar, 2015). However, all these approaches are ostensibly focused on
life periods where the general social context is relatively stable. In this
article, we present our conceptualization of the family in times of
unprecedented social upheaval. Specifically, our framework seeks to characterize
family dynamics as a suitable lens to analyze health inequalities in the face of
a virulent public health crisis.
